---
title: "ps3.qmd"
format: html
editor: visual
---

```{r}
#1.1 Load the tidyverse, glmnet, and knitr packages. Then load the COMPAS dataset from data/compas_clean.csv.

install.packages("glmnet")

library(tidyverse)
library(glmnet)
library(knitr)

compas <- read_csv("/Users/ubpham/gov51/gov51-ps3/data/raw/compas_clean.csv")
```

```{r}
#1.2 How many observations and variables are in the dataset? What is the unit of observation?

head(compas)

nrow(compas)
ncol(compas)
```

There are 7214 observations and 17 variables.

Based on head (see the first few rows of the data set), each row appears to be one person (based on their age, gender, recidivism, race).

```{r}
#1.3 Examine the outcome variable The variable recidivism equals 1 if the person was rearrested within two years and 0 otherwise.

# (a) What is the overall recidivism rate in the data?

mean(compas$recidivism)

# (b) Is this outcome continuous or binary? Why does this matter for how we interpret predictions from a linear model?
```

Overall recidivism rate 45.1% (0.45065, recidivism is stored as a binary, so take mean of 1 and 0).

The outcome is binary. Recidivism is 1 or 0; 1 if the other person WAS arrested within two years and 0 if the person doesn't meet that condition.

This matters because the linear model OLS may predict values below 0 or above 1; however, since recidivism is binary, this may fall outside the valid probability range (probability from 0 to 1). The linear model predictions, however, should be interpreted to predicting the probability of recidivism; though, OLS is not restricted when predicting for binary variables. We should be aware the linear model may predict unrealistic probabilities outside the range of 0 to 1 (not valid probability).

```{r}
#1.4 Create a summary statistics table showing the mean and standard deviation of all variables in the dataset. Briefly note which variables have the most variation and which have the least.

summary_table <- compas %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  group_by(variable) %>%
  summarise(mean = mean(value), sd = sd(value))

kable(summary_table, digits = 3)
```

The variable age has the most variation (standard deviation of 11.889), followed by priors_count (standard deviation of 4.883).

Native American has the least variation (standard deviation of 0.050), followed by the Asian variable with the second to least variation (standard deviation of 0.066).

```{r}
#1.5 Create a bar chart showing the recidivism rate by number of prior offenses (priors_count).

# Group priors_count into categories: 0, 1–2, 3–5, 6–10, and 11+. What pattern do you see?

compas <- compas %>%
  mutate(priors_cat = case_when(
    priors_count == 0 ~ "0",
    priors_count %in% 1:2 ~ "1-2",
    priors_count %in% 3:5 ~ "3-5",
    priors_count %in% 6:10 ~ "6-10",
    priors_count >= 11 ~ "11+"
  )) %>%
  mutate(priors_cat = factor(priors_cat,
                             levels = c("0", "1-2", "3-5", "6-10", "11+")))

recid_by_priors <- compas %>%
  group_by(priors_cat) %>%
  summarise(recid_rate = mean(recidivism),
            n = n())

kable(recid_by_priors, digits = 3)

ggplot(recid_by_priors, aes(x = priors_cat, y = recid_rate)) +
  geom_col() +
  labs(title = "Recidivism Rate by Prior Offenses Category",

T
       x = "Prior Offenses Category",
       y = "Recidivism Rate") +
  theme_minimal()
```

The pattern that I see is that as the number of prior offenses increases, the recidivism rate appears to increase as well. Those with 0 prior offenses have the lowest recidivism rate, while those with 11+ prior offenses have the highest recidivism rate, with a consistent increase in recidivism rate as the prior offenses category increase.

```{r}
#2.1 Set your seed to 51 for reproducibility. Split the data into a training set (80 ) and a test set (20%)

set.seed(51)
n <- nrow(compas)
train_idx <- sample(1:n, size = floor(0.8 * n))
train <- compas[train_idx, ]
test <- compas[-train_idx, ]

#How many observations are in each set?

nrow(train)
nrow(test)
```

```{r}
#2.2 Fit an OLS regression of recidivism on all other variables in the training data using lm(recidivism ~ ., data = train). This gives you a model with 16 predictors.

train <- train %>% select(-priors_cat)
test <- test %>% select(-priors_cat)

ols_simple <- lm(recidivism ~ ., data = train)
summary(ols_simple)

#(a) Which predictors are statistically significant at the 5% level?
#(b) Interpret the coefficient on priors_count in one sentence.
```

The predictors that are statistically significant at the 5% level have t-value less than 1.96 or greater than 1.96. This includes age, female, priors_count, felony, age between 25 and 45, has_juv_misd.

Priors_count coefficient is 0.029532. This means, according to the OLS model, for every 1 count increase in a person's prior counts of crime, their rate of recidivism (percentage) will increase by 2.9532 percentage points.

```{r}
#2.3 Root Mean Squared Error (RMSE) measures average prediction error:

rmse <- function(actual, predicted) sqrt(mean((actual - predicted)^2))

# (a) Compute the in-sample RMSE (predictions on the training data).

pred_train_simple <- predict(ols_simple, newdata = train)
rmse_train_simple <- rmse(train$recidivism, pred_train_simple)
rmse_train_simple

# (b) Compute the out-of-sample RMSE (predictions on the test data).

pred_test_simple <- predict(ols_simple, newdata = test)
rmse_test_simple <- rmse(test$recidivism, pred_test_simple)
rmse_test_simple

# (c) How large is the gap between in-sample and out-of-sample RMSE? Is OLS overfitting with 16 predictors?

rmse_test_simple - rmse_train_simple# (a) Compute the in-sample RMSE (predictions on the training data).
# (a) Compute the in-sample RMSE (predictions on the training data).


#Hint: Use predict(model, newdata = ...) to generate predictions, then compute RMSEmanually.
```

```{r}
#2.4 A single train/test split depends on which observations happened to land in each group. Crossvalidation averages over multiple splits to give a more stable estimate of out-of-sample performance. 
# Implement 5-fold cross-validation for the simple OLS model:
# 1. Randomly assign each training observation to one of 5 folds.
# 2. For each fold 𝑘 = 1, … , 5: fit OLS on the other 4 folds, predict fold 𝑘, compute RMSE.
# 3. Average the 5 RMSE values.

# (a) What is the 5-fold CV estimate of RMSE? How does it compare to your single-split out-of-sample RMSE?
# (b) In your own words, why is cross-validation more reliable than a single train/test split?
# (c) When we use cv.glmnet() later, it does this same process internally to choose the penalty parameter 𝜆. Why is cross-validation the right tool for choosing 𝜆?
```

```{r}
#3.1 The formula below creates all pairwise interactions plus squared terms for the continuous variables. This expands our feature set from 16 to about 141 predictors.

# (a) How many predictors does this expanded model have?
# (b) What is the ratio of training observations to predictors? (Compare to the simple model.
```

```{r}
#3.2 Fit OLS on the expanded feature set: ols_full <- lm(f_full, data = train).

# (a) Compute the in-sample RMSE for this model.
#(b) Compute the out-of-sample RMSE.
# (c) How does the in-sample/out-of-sample gap compare to the simple 19-predictor OLS? What happened and why?
```

```{r}
#3.3 In 2–3 sentences, explain why adding 180 interaction terms made the model’s in-sample fit better but its out-of-sample predictions worse. What is this phenomenon called?
```

```{r}
#4.1 Use cv.glmnet() with alpha = 0 to fit Ridge regression on the expanded training data (the X_train matrix with ~141 features). Use 10-fold cross-validation (the default) to select the optimal 𝜆.

# (a) What is the optimal 𝜆 selected by cross-validation? (Use lambda.min.)
# (b) Plot the cross-validation curve using plot(cv_ridge). What does this plot show?
```

```{r}
#4.2 Compute the out-of-sample RMSE for Ridge. Compare to both the simple OLS (16 vars) and the kitchen-sink OLS (~141 vars). What happened?
```

```{r}
#4.3 Extract the Ridge coefficients using coef(cv_ridge, s = "lambda.min"). Did Ridge eliminate any variables? Why or why not?
```

```{r}
#5.1 Use cv.glmnet() with alpha = 1 on the expanded training data. Remember to set your seed first for reproducibility.

#(a) What is the optimal 𝜆?
#(b) Plot the cross-validation curve.
```

```{r}
#5.2 Extract the LASSO coefficients at lambda.min.

#(a) How many of the ~141 variables survived (non-zero coefficient)? How many were eliminated?
#(b) List a few of the surviving interaction terms. Do they make substantive sense?
# (c) Baker (2025) argues that LASSO reduces “subjective discretion” in model building. How does what you see here illustrate that argument?
```

```{r}
#5.3 Compute the out-of-sample RMSE for LASSO. How does it compare to OLS (both vers
```

```{r}
#6.1 Fit an Elastic Net with alpha = 0.5 on the expanded training data. Remember to set your seed first. Compute its out-of-sample RMSE and count how many variables it retains.
```

```{r}
#6.2 Fill in this table with your results:

#(a)Which model has the lowest out-of-sample RMSE?
# (b) Which model has the best in-sample fit? Is that the same model?
# (c) In 2–3 sentences, explain what this table teaches about the bias-variance tradeoff.
```

```{r}
#6.3 In Baker (2025), two experts in the Halliburton securities case selected different peer firms and reached different conclusions about whether a stock drop was statistically significant.

#(a) How did LASSO help narrow the disagreement between dueling experts?
# (b) Based on your experience with LASSO in this problem set, what aspect of the traditional OLS approach gives the analyst the most room for manipulation?
```

```{r}
#6.4 We used these models to predict recidivism. Suppose instead you wanted to know whether priors_count causes higher recidivism.

#(a) Would the LASSO coefficient on priors_count answer this causal question? Why or why not?
#(b) What is the difference between using regression for prediction versus using it for causal inference?
```

```{r}
#6.5 Algorithms like COMPAS are used in real courtrooms to inform bail and sentencing decisions.

#(a) Using your simple OLS model (16 vars), compute the mean predicted recidivism probability separately for Black defendants and white defendants in the test set. Is there a difference?

#Hint: White is the reference category (all race dummies equal 0). You can identify white defendants with black == 0 & hispanic == 0 & asian == 0 & other_race == 0& native_american == 0. 

# (b) Now compute the actual recidivism rate for each group. Compare the predictions to the actual rates. 
# (c) In 2–3 sentences: What are the tradeoffs of using prediction algorithms in criminal justice? Reference something specific from your analysis.
```
