---
title: "ps1.qmd"
format: html
editor: visual
---

```{r}
# 1.1.1 Load packages and data

library(tidyverse)
library(knitr)
gay <- read_csv("/Users/ubpham/gov51/gov51-ps2/data/raw/gay.csv")
```

```{r}
# Q1.1 How many total observations are in gay.csv? How many data rows to compare?
nrow(gay)
# How many unique values does study take? How many different studies in the data set
length(unique(gay$study))
# List all the treatment conditions in the treatment variable.
unique (gay$treatment)
```

```{r}
# Q1.2 Filter to Study 1, Wave 1 only. Calculate the mean of ssm (same-sex marriage attitude, scored 1–5) for each treatment group using group_by() and summarize(). Which treatment group shows the highest average support?

# WITHIN GAY, filter Study 1 Wave 1, Froup By by treatment group.
# Ssm = same-sex marraige attitude variable, which I'm taking the mean of. 
gay %>% 
  filter(study == 1, wave == 1) %>% 
  group_by(treatment) %>% 
  summarise(mean_same_sex_attitude = mean(ssm, na.rm = TRUE), n())
```

Recycling Script by Gay Canvasser had the highest mean at 3.13. In addition, all means are close to 3 at the baseline, indicating that the groups are roughly similar which reflects random assignment.

```{r}
# Q1.3 Still using Study 1, Wave 1, calculate the standard deviation of ssm within each treatment group. Are the standard deviations roughly similar across groups? Why is this important in a randomized experiment? (2–3 sentences)
gay %>% 
  filter(study == 1, wave == 1) %>% 
  group_by(treatment) %>% 
  summarise(sd_same_sex_attitude = sd(ssm, na.rm = TRUE))
```

Interpretation: The standard deviations appear roughly similar across the groups and within the range of 1.66 to 1.70. This means that each treatment group has roughly the same spread of sample data. This is important for randomization (when randomly selecting a sample) because in order to draw conclusions, each group needs to be roughly similar in factors beside the treatment group.

```{r}
# Q1.4 Create a bar plot of the mean ssm by treatment group (Study 1, Wave 1) using ggplot() with geom_col(). Hint: First create a summary data frame with the means, then plot it.

summary_mean_ssm <- gay %>% 
  filter(study == 1, wave == 1) %>% 
  group_by(treatment) %>% 
  summarise (mean_same_sex_attitude = mean(ssm, na.rm = TRUE))

ggplot(summary_mean_ssm, aes(x = treatment, y = mean_same_sex_attitude)) +
  geom_col(fill = "steelblue") +
  labs (x = "Treatment Group", 
        y = "Mean Same Sex Marriage Attitude", 
        title = "Mean SSM by Treatment Group") +
        theme(axis.text.x = element_text(angle = 30, hjust = 1, size = 7))
```

```{r}
# Q1.5 A well-run randomized experiment should produce groups that look similar at baseline. Filter gay.csv to Study 1, Wave 1. Compare the “Same-Sex Marriage Script by Gay Canvasser” group and the “No Contact” group:

# • Calculate the mean and SD of ssm for each group

gay_samescript_gaycanvas <- gay %>% filter(study == 1, wave == 1, 
                              treatment == "Same-Sex Marriage Script by Gay Canvasser")
mean_samescript_gaycanvas <- mean(gay_samescript_gaycanvas$ssm, na.rm = TRUE)
sd_samescript_gaycanvas <- sd(gay_samescript_gaycanvas$ssm, na.rm = TRUE)

mean_samescript_gaycanvas
sd_samescript_gaycanvas

gay_nocontact <- gay %>% filter(study == 1, wave == 1, 
                             treatment == "No Contact")
mean_nocontact <- mean(gay_nocontact$ssm, na.rm = TRUE)
sd_nocontact <- sd(gay_nocontact$ssm, na.rm = TRUE)

mean_nocontact
sd_nocontact

# • Compute the standardized difference in means.
# Treated = gay canvasser, Nontreated = no contact.

std_mean <- (mean_samescript_gaycanvas - mean_nocontact) / sqrt( (sd_samescript_gaycanvas^2 + sd_nocontact^2)/2)

std_mean

# A standardized difference below 0.25 in absolute value suggests good balance. Do these groups look similar at baseline? (2–3 sentences)
```

The standard difference between the treatment group (same-sex script with a gay canvasser) and the control group (no contact group) is 0.01, which is below 0.25. Therefore, the standard difference suggest a good balance for both the means (same-sex marriage attitude) and standard deviation of the data at baseline (Wave 1 is before any contact). More broadly, since we're able to quantify the extent of the difference between the means (first compared in Q1.4), it shows that our random samples for each groups is roughly similar and not largely influenced by differences in covariates.

```{r}
# Q1.6 Now look at the outcomes. Filter to Study 1, Wave 2 only. Calculate the mean ssm for the gay canvasser group and the No Contact group. What is the difference in means (gay canvasser minus no contact)? Interpret: since the groups were balanced at baseline, what does a difference at Wave 2 suggest? (2–3 sentences)

gay_samescript_gaycanvas <- gay %>% filter(study == 1, wave == 2, 
                              treatment == "Same-Sex Marriage Script by Gay Canvasser")
mean_samescript_gaycanvas <- mean(gay_samescript_gaycanvas$ssm, na.rm = TRUE)

mean_samescript_gaycanvas

gay_nocontact <- gay %>% filter(study == 1, wave == 2, 
                             treatment == "No Contact")
mean_nocontact <- mean(gay_nocontact$ssm, na.rm = TRUE)

mean_nocontact

mean_wave2_difference = mean_samescript_gaycanvas - mean_nocontact

mean_wave2_difference
```

The difference in means between the treatment group (same sex script with gay canvassers) and control group (no contact) is 0.10. Since the groups were balanced at baseline (according to our calculation for the standard difference in means), this suggests that the treatment group caused a small, positive increase in same sex marriage attitudes and indicates a possible treatment effect after just one wave.

```{r}
# Q2.1  Load gayreshaped.csv and ccap2012.csv. Report the number of observations and column names for each dataset. In gayreshaped.csv, the variables therm1 through therm4 represent feeling thermometer scores (0–100) toward gay men and lesbians, measured at each survey wave.

gay_r <- read_csv("/Users/ubpham/gov51/gov51-ps2/data/raw/gayreshaped.csv")
ccap <- read_csv("/Users/ubpham/gov51/gov51-ps2/data/raw/ccap2012.csv")

nrow(gay_r)
colnames(gay_r)

nrow(ccap)
colnames(ccap)
```

```{r}
# Q2.2 Extract the baseline feeling thermometer (therm1) for Study 1 from gayreshaped.csv. Calculate its mean, median, and standard deviation. 

study1_gayr <- gay_r %>% 
  filter(study == 1,) %>% 
  pull(therm1)

mean_study1 <- mean(study1_gayr, na.rm = TRUE)
med_study1 <- median(study1_gayr, na.rm = TRUE)
sd_study1 <- sd(study1_gayr, na.rm = TRUE)

# Then do the same for gaytherm from ccap2012.csv (remove NAs). 

ccap_gaytherm <- ccap %>% 
  pull(gaytherm)

mean_ccap <- mean(ccap_gaytherm, na.rm = TRUE)
med_ccap  <- median(ccap_gaytherm, na.rm = TRUE)
sd_ccap   <- sd(ccap_gaytherm, na.rm = TRUE)

# Report both sets of statistics side by side. What do you notice? (1–2 sentences)

summary_stats <- tibble(
  Variable = c(
    "Gay Study 1",
    "CCAP 2012"
  ),
  Mean = c(
    mean_study1,
    mean_ccap
  ),
  Median = c(
    med_study1,
    med_ccap
  ),
  StdDev = c(
    sd_study1,
    sd_ccap
  )
)
summary_stats |>
  kable(
    col.names = c("Variable","Mean","Median","Std_Dev"),
    digits = 3,
    caption = "Table 1: Summary Statistics for Study 1 via Gay and GayTherm via CCAP 2012",
    )
```

Overall, the statistics between Study 1 of Gay Reshaped and CCAP 2012 appear to be very similar. Between the Lacour baseline and the CCAP 2012 (which is actual data), the mean same sex marriage attitude differs by 0.4, the medians by 2, and standard_deviations by 0.8.

```{r}
# Q2.3  Create side-by-side histograms comparing the LaCour Study 1 baseline (therm1) and the CCAP data (gaytherm). Use the same bin breaks for both: breaks = seq(0, 100, by = 5). 

study1 <- gay_r %>%
  filter(study == 1) %>%
  select(category = therm1) %>%
  mutate(source = "Study 1")

ccap_data <- ccap %>%
  select(category = gaytherm) %>%
  mutate(source = "CCAP")

combined <- bind_rows(study1, ccap_data) %>%
  filter(!is.na(category))

ggplot(combined, aes(x = category)) +
  geom_histogram(breaks = seq(0, 100, by = 5)) +
  facet_wrap(~source, scales = "free_y") +
  labs(
    x = "Same Sex Attitude Thermometer Measure", 
    y = "Respondent Count",
    title = "Same-Sex Feeling Thermometers from Reshaped & CCAP"
  )

# Do these look like they came from different populations? (2–3 sentences)
```

No, the data does not appear to be from different populations; instead, the shapes of both distributions as shown by the histogram appear extremely similar. There are peaks in both graphs near the feeling thermometer of 0, 50, and 100. In both histograms, the clustered data is higher at 100 score compared to 0 score, as well as the peak before 50 score being larger than the peak after 50 score. Overall, the growth of respondent counts also gradually increases between 0 to 50 score and then 50 to 100 score in both histograms.

```{r}
# Q2.4 For both the Study 1 and Study 2 control groups (“No Contact”), calculate:

  # (a) The correlation between therm1 and therm2 (use cor() with use = "complete.obs")

therm1_study1 <- gay_r %>% 
  filter(study == 1, treatment == "No Contact") %>% 
  pull(therm1)

therm2_study1 <- gay_r %>% 
  filter(study == 1, treatment == "No Contact") %>% 
  pull(therm2)

cor_study1 = cor(therm1_study1, therm2_study1, use = "complete.obs")
cor_study1

therm1_study2 <- gay_r %>% 
  filter(study == 2, treatment == "No Contact") %>% 
  pull(therm1)

therm2_study2 <- gay_r %>% 
  filter(study == 2, treatment == "No Contact") %>% 
  pull(therm2)

cor_study2 = cor(therm1_study2, therm2_study2, use = "complete.obs")
cor_study2

  # (b) The standard deviation of the within-person change (therm2 - therm1)

sd_study1 = sd(therm2_study1-therm1_study1, na.rm=TRUE)
sd_study1
sd_study2 = sd(therm2_study2-therm1_study2, na.rm=TRUE)
sd_study2

# Present your results side by side. A re-test correlation of 0.95–0.97 is typical for real feeling thermometer data. Which study shows a more plausible pattern, and why? (3–4 sentences)
```

Study 2 (a treatment group) has a re-test correlation of 0.97 (between therm 1 intiial test and therm 2 follow up test) , which mean the correlation falls within a typical range for a real feeling thermometer data. However, Study 1's (a control group) correlation of 0.998 (between therm 1 intiial test and therm 2 follow up test) falls outside the re-test correlation typical range. In fact, Study 1 has a correlation stronger (greater than) the typical correlation range. For Study 1, even for a control group, such a strong correlation means that respondents were giving the same answers in both rounds therm1 and round therm2, which is unlikely due to other randomization and variation.

```{r}
# Q2.5 Create two scatter plots side by side:
# (a) Study 1 control group: therm1 (x-axis) vs therm2 (y-axis), with a 45-degree line

gay_r %>%
  filter(study == 1, treatment == "No Contact") %>%
  ggplot(aes(therm1, therm2)) +
  geom_point(alpha = 0.3, na.rm = TRUE) +
  geom_abline(slope = 1, intercept = 0) + 
  labs(
    title = "Study 1 — The Control Group Baseline and Follow-Up",
    x = "Baseline Round 1 Thermometer (therm1)",
    y = "Follow-Up Round 2 Thermometer (therm2)"
  )
  
# (b) Study 2 control group: therm1 (x-axis) vs therm2 (y-axis), with a 45-degree line
  
gay_r %>%
  filter(study == 2, treatment == "No Contact") %>%
  ggplot(aes(therm1, therm2)) +
  geom_point(alpha = 0.3, na.rm = TRUE) +
  geom_abline(slope = 1, intercept = 0) +
  labs(
    title = "Study 2 — A Treatment Group Baseline and Follow-Up",
    x = "Baseline Round 1 Thermometer (therm1)",
    y = "Follow-Up Round 2Thermometer (therm2)"
  )

# Describe the differences. Which looks more like real longitudinal data, and why? (3–4 sentences)
```

Study 2 looks more like real longitudinal data. Both have clusters on the round 1 thermometer score of 50 and 100 (on the x-axis), meaning there are clusters of data points heaping on the corresponding score. While both distributions generally follow the trend of a positive linear relationship between the thermometer score of the baseline round and follow-up round, in Study 1, the data clusters strongly directly on the 45 degree line (extremely low residuals). On the other hand, in Study 2, the data follows the general trend of a positive linear relationship but the data has a higher spread from the 45 degree line as well as greater residuals from the 45 degree line. Therefore, Study 2 looks more like real longitudinal by reflecting the variability within data and scattering around the 45 degree line; true data is not as strongly correlated as displayed by Study 1 due to variability and randomization within sampling.

```{r}
# Q2.6 You’ve now found three irregularities:
# 1. The baseline distributions match an existing national survey
# 2. The re-test correlations are implausibly high (Study 1)
# 3. The within-person changes are implausibly small (Study 1)

# In 4–5 sentences, explain why these three findings together suggest the data were fabricated rather than collected in a real experiment. Use specific numbers from your analysis. Why would a fabricator produce data with these properties?
```

In Q2.3, we established that the baseline distributions for the reshaped gay distribution and the CCAP existing national survey displayed strong similarities. There are peaks in both graphs near the feeling thermometer of 0, 50, and 100. In both histograms, the clustered data is higher at 100 score compared to 0 score, as well as the peak before 50 score being larger than the peak after 50 score. Overall, the growth of respondent counts also gradually increases between 0 to 50 score and then 50 to 100 score in both histograms.

Therefore, the similar baseline distributions suggest that the data was not randomly and independently sampled in the reshaped gay distribution; instead the data was likely taken from an existing national survey data.

Instead, the baseline distribution between the existing national survey and reshaped gay distribution should share weaker similarities since they took different samples.

In Q2.4, Study 1's (a control group) correlation of 0.998 (between initial therm1 and retest therm2) falls outside the re-test correlation typical range of 0.95-0.97. On the other hand, Study 2 has a correlation of 0.97 between initial therm 1 and retest therm 2, which falls within the typical re-test correlation typical range. Such a strong correlation for Study 2 indicates that respondents in this sample would have responded with the same thermometer score two times in a row, which is extremely unlikely to be such strongly consistent due to variability between responses.

Such precision and strong correlation therefore suggests the data was not authentically generated, since it does not reflect the true variability that comes with sampling real people and slightly varying thermometer responses between the initial therm1 and follow-up therm2.

Instead, the correlation relating initial therm1 and retest therm 2 should fall within the typical re-test correlation range with a slightly weaker relationship between initial and subsequent thermometer score relationships.

In Q2.5, Study 1 had an extremely strong correlation (low residuals) that clustered closely to the 45 degree line. Study 2 also followed the strong positive linear relationship of the 45 degree line, but the data points were more scattered and had higher residuals from the true 45 degree line.

Such a strong correlation in Study 1 would signal falsified data since it does not reflect the true variability and "noise" that would come with sampling (the relationship between therm1 thermometer score and therm2 thermometer score would not be as strongly consistent with a consistent trend).

Instead, the data points should be more scattered and less clustered to the 45 degree line. The distribution can still follow the general positive linear relationship trend, but showing a less strong correlation.

A fabricator would produce data with these properties, as reflected in Study 1, because they would want to achieve certain "perfect" conclusions without having to account for variability that occurs within sampling.

```{r}
# Q3.1 Filter gayreshaped.csv to Study 2. How many observations? Calculate the mean and SD of therm1 (baseline) for the treatment group and the control group separately.

study2 <- gay_r %>% filter(study == 2)

study2 %>% 
   summarise (n=n())

study2 %>% 
  group_by(treatment) %>% 
  summarise(
    n=n(),
    mean_therm1 = mean(therm1, na.rm = TRUE),
    sd_therm1 = sd(therm1, na.rm = TRUE),
    mean_therm2 = mean(therm2, na.rm = TRUE),
    sd_therm2 = sd(therm2, na.rm = TRUE))
```

```{r}
# Q3.2 Create side-by-side histograms of Study 1 baseline (therm1) and Study 2 baseline (therm1), using the same bins as before. How do they compare visually? (2–3 sentences)

study1 <- gay_r %>%
  filter(study == 1)

study2 <- gay_r %>%
  filter(study == 2) 

combined <- bind_rows(
  tibble(source = "Study 1 Baseline", therm = study1$therm1),
  tibble(source = "Study 2 Baseline", therm = study2$therm1)
) %>% filter (!is.na(therm))

ggplot(combined, aes(x = therm))+
  geom_histogram(breaks = seq(0, 100, by = 5)) +
  facet_wrap(~source, scales = "free_y") +
  labs(
    x = "Same Sex Attitude Thermometer Measure", 
    y = "Respondent Count",
    title = "Feeling Thermometers Baseline from Treatment & Control"
  )
```

Both the baseline distribution of Study 1 (control) and Study 2 (treatment) are roughly similar in distribution shape and peaks. The shapes are broadly consistent between both baseline groups, indicating that the sampling groups for both the treatment and control group are roughly similar even after randomization. Sample 1 has a larger sample size. However, Study 2 also shows meaningful variation.

```{r}
# Q3.3 Filter gayreshaped.csv to Study 2, keeping only the “Same-Sex Marriage Script by Gay Canvasser” and “No Contact” groups. Check balance at baseline: calculate the mean and SD of therm1 for each group and the standardized difference in means (same formula as Q1.5). Do the groups look balanced? (1–2 sentences)

study2_treat <- study2 %>%
  filter(treatment == "Same-Sex Marriage Script by Gay Canvasser")

mean_samescript_gaycanvas <- mean(study2_treat$therm1, na.rm = TRUE)
mean_samescript_gaycanvas
sd_samescript_gaycanvas <- sd(study2_treat$therm1, na.rm = TRUE)
sd_samescript_gaycanvas

study2_control <- study2 %>%
  filter(treatment == "No Contact")

mean_nocontact <- mean(study2_control$therm1, na.rm = TRUE)
sd_nocontact <- sd(study2_control$therm1, na.rm = TRUE)
mean_nocontact
sd_nocontact

std_mean <- (mean_samescript_gaycanvas - mean_nocontact) / sqrt( (sd_samescript_gaycanvas^2 + sd_nocontact^2)/2)

std_mean
```

The groups look largely balanced due to the standardized difference in means 0.05128 being less than 0.25. Therefore, the standard difference suggest a good balance for both the means (same-sex marriage attitude) and standard deviation of the data at baseline (Therm 1 is before any contact).

```{r}
# Q3.4 Now compare outcomes. Calculate the mean therm2 (post-conversation) for each group. What is the difference in post-period means (gay canvasser minus no contact)?

study2_treat <- study2 %>%
  filter(treatment == "Same-Sex Marriage Script by Gay Canvasser")

mean_samescript_gaycanvas <- mean(study2_treat$therm2, na.rm = TRUE)
mean_samescript_gaycanvas

study2_control <- study2 %>%
  filter(treatment == "No Contact")

mean_nocontact <- mean(study2_control$therm2, na.rm = TRUE)
mean_nocontact

difference_means = mean_samescript_gaycanvas - mean_nocontact
difference_means
```

```{r}
# Q3.5 Calculate the standard error of the difference in post-period means from Q3.4.2 is the variance of therm2 and 𝑛 is the sample size for each group. 

study2_treat <- study2 %>%
  filter(treatment == "Same-Sex Marriage Script by Gay Canvasser")
  n_treatment = sum(!is.na(study2_treat$therm2))

var_samescript_gaycanvas <- var(study2_treat$therm2, na.rm = TRUE)

study2_control <- study2 %>%
  filter(treatment == "No Contact")
  n_control = sum(!is.na(study2_control$therm2))

var_nocontact <- var(study2_control$therm2, na.rm = TRUE)

standard_error = sqrt (
  (var_samescript_gaycanvas/ n_treatment) + 
  (var_nocontact / n_control)
  )

standard_error

# Then construct a 95% confidence interval: ( ̄𝑥1 − ̄𝑥2) ± 1.96 × 𝑆𝐸Does the CI include zero? What does this tell you about whether the groups differ after the conversations?

lower_bound = difference_means - 1.96 * standard_error
upper_bound = difference_means + 1.96 * standard_error
lower_bound
upper_bound
```

The confidence interval from 1.133 to 6.00 of the difference of means does not capture 0 (which means we can reject the null hypothesis that there is no difference between mean of control and mean of treatment). We are 95% confident that the interval of 1.133 to 6.00 captures the true difference between the mean of the treatment and mean control group; we are 95% confident that there is a difference between the groups after the conversations.

```{r}
# Q3.6 For the Study 2 control group, calculate the correlation between therm1 and each subsequent wave: therm2, therm3, and therm4. 
# Use cor() with use = "complete.obs". Do the correlations decay over time (i.e., do more distant waves correlate less with baseline)? Wha does this pattern tell you about the stability of real survey measurements? (2–3 sentences)

study2_control <- study2 %>%
  filter(treatment == "No Contact")
  cor1= cor(study2_control$therm1, study2_control$therm2, use = "complete.obs")
  cor2= cor(study2_control$therm1, study2_control$therm3, use = "complete.obs")
  cor3= cor(study2_control$therm1, study2_control$therm4, use = "complete.obs")
  cor1
  cor2
  cor3
```

The correlations do not decay over time; the correlations do fluctuate slightly (decrease for correlation of therm3 and therm1, increase for correlation of therm2 and therm1). This pattern tells me that there is a strong stability of real survey measurements about the effect of the treatment conversations on people's attitude thermometer. Per 2.4, all three correlations are relatively close to the interval of 0.95 to 0.97, which is a normal range for correlation of feeling thermometers and reflects real world data.

```{r}
# Q3.7 You’ve now analyzed fabricated data (Study 1), caught the fraud, and examined more realistic data (Study 2). In 5–6 sentences, discuss:
# • What makes Study 2’s data different from Study 1’s data? Be specific about which statistics differ and by how much.
# • Why is it important that real treatment effects are smaller and messier than fabricated ones?
# • What role did simple measurement tools (mean, SD, histograms, correlations) play in detecting the fraud?
```

Scatterplot

Correlation

Standard Deviation

```{r}
# Q4.1 For the Study 1 control group, fit a linear regression predicting therm2 from therm1:
# therm2𝑖 = 𝛽0 + 𝛽1 × therm1𝑖 + 𝜖𝑖
# Store the result as fit_s1 and display the summary. Report the slope (𝛽̂ )1 the R-squared, and the residual standard error
# Use lm(therm2 ~ therm1, data = ...) and summary().

fit_study1 <- lm (therm2 ~ therm1, data = gay_r %>% 
      filter(study == 1, treatment == "No Contact"))

summary(fit_study1)
```

The slope is 0.992. Every 1 point increase on the attitude thermometer score at baseline nearly predicts the attitude thermometer score on the first-follow up.

The R-squared is 0.9952. 99.52% of the variation in attitude thermometer scores on the first-follow up is accounted for by baseline attitude thermometer scores. It explains almost all the variation.

The residual standard error is 1.976. On average, an actual attitude thermometer score on the first-follow up is 1.976 points away from the predicted attitude thermometer score.

```{r}
# Q4.2 Repeat Q4.1 for the Study 2 control group. Store as fit_s2. Report the same three quantities: slope, R-squared, and residual standard error.

fit_s2 <- lm(therm2 ~ therm1,
             data = gay_r %>% filter(study == 2, treatment == "No Contact"))
summary(fit_s2)

# Compare the two models. In fabricated data, we’d expect 𝛽̂1 ≈ 1 and 𝑅2 ≈ 1 because the “followup” data is just a copy of the baseline with tiny noise added. In real data, we’d expect regression to the mean: 𝛽̂ < 1 and lower R^2. Which pattern do you see? (3-4 sentences)
```

The slope is 0.969. Every 1 point increase on the attitude thermometer score at baseline predicts a 0.969 increase in attitude thermometer score on the first-follow up.

The R-squared is 0.9476. 94.76% of the variation in attitude thermometer scores on the first-follow up is accounted for by baseline attitude thermometer scores. It explains almost all the variation, but less than Study 1.

The residual standard error is 6.547. On average, an actual attitude thermometer score on the first-follow up is 6.457 points away from the predicted attitude thermometer score. This is a larger residual compared to Study 1 (less accurate predictions).

COMPARISON: I see the pattern of fabricated data in Study 1. The slope (beta-1) of 0.992 is nearly equal to 1 and the R-squared of 0.995 is nearly equal to 1 as well, which indicates a strong correlation and also almost all the variation in attitude thermometer score on follow-up being accounted for, which is indicative of fabricated data.

For Study 2, the slope (beta-1) of 0.969 is less than 1 compared to Study 1's slope (0.992) and the R-squared of 0.948 is also less than 1 and Study 1's R-squared of 0.995. The weaker correlation and the less variation in attitude thermometer score on follow-up accounted for is indicate of real data compared to Study 1.

```{r}
# Q4.3 Now use regression to compare the two groups in Study 2. Create a binary variable treated that equals 1 for the “Same-Sex Marriage Script by Gay Canvasser” group and 0 for the “No Contact” group. 

# Then fit: therm2𝑖 = 𝛽0 + 𝛽1 × treated𝑖 + 𝜖𝑖
# Report 𝛽̂1, its standard error, t-statistic, and p-value.
# Compare 𝛽̂1to the difference in means you calculated in Q3.4 — what do you notice? 
# What does 𝛽̂0estimate?
# Is the difference between groups statistically significant at 𝛼 = 0.05?

d2 <- gay_r %>%
  filter(study == 2,
         treatment %in% c("Same-Sex Marriage Script by Gay Canvasser", "No Contact")) %>%
  mutate(treated = if_else(treatment ==
           "Same-Sex Marriage Script by Gay Canvasser", 1, 0))

summary(lm(therm2 ~ treated, data = d2))

# Note: Filter to only the gay canvasser and no contact groups before running the regression

```

The beta 1 hat is 3.57. It is the same as the difference of the treatment therm2 mean (follow-up) and control therm2 mean (follow-up), which is also 3.57.

The standard error is 1.24.

Beta zero is 57.63. This represent the y-intercept. In other words, when the control (control group, beta hat 1) thermometer follow-up score is 0 points, the treatment thermometer follow-up score is 57.63 points.

The t-score is 2.87.

The p-value is 0.004. Since the p-value of 0.004 is less than alpha of 0.05, there is a statistically significant difference between the treatment and control groups of follow up attitude thermometer scores.

```{r}
# Q4.4 Add therm1 as a control variable:
# therm2𝑖 = 𝛽0 + 𝛽1 × treated𝑖 + 𝛽2 × therm1𝑖 + 𝜖𝑖

fit2 <- lm(therm2 ~ treated + therm1, data = d2)
summary(fit2)

# Report the coefficient on treated, its standard error, t-statistic, and p-value. 

# Compare this to your results from Q4.3:
# (a) Did the estimate of 𝛽̂1 change much?
# (b) Did the standard error change? In which direction?
# (c) Why does controlling for baseline attitudes improve precision? (2–3 sentences)
```

The new coefficient on treated is 2.46. The previous coefficient on treated was 3.57. This is a moderate decrease.

The standard error is now 0.34. The previous standard error was 1.24. This is approximately a 73% decrease, so this is a significant decrease change.

Adding the therm1 control variable (baseline attitude thermometer score) explains more of the variation between follow-up attitude thermometer scores between treatment and control group. Therefore, controlling for the baseline attitudes improves precision.

```{r}
# Q4.5 Compare the R-squared values from Q4.3 and Q.4. Why does adding therm1increase the R-squared so dramatically? What does this tell you about how much of the variation in follow-up attitudes is explained by where people started versus the treatment?

# Finally, use predict() to generate predicted values from your Q4.4 model for two hypothetical respondents:
# • Person A: treated = 1, therm1 = 50 (moderate baseline)
# • Person B: treated = 0, therm1 = 50 (same baseline, no treatment)

person_a = predict(fit2, newdata = data.frame (treated = 1, therm1 = 50))

person_b = predict(fit2, newdata = data.frame (treated = 0, therm1 = 50))

person_a
person_b

pred_difference = person_a-person_b
pred_difference

# What is the predicted difference in their Wave 2 attitudes?
```

Q4.3 R squared was 0.0039.

Q4.4 R squared is 0.925.

The R-squared increased dramatically, meaning the linear model in 4.4 accounts for 92.5% of the variation in attitude thermometer scores in the treatment follow-up, compared to only 3.9% of the variation accounted for under the 4.3 linear model. This tells me that nearly 90% of the variation (a large significant percentage of the total variation) is accounted for by people's baseline attitude temperature scores.

The predicted difference is 2.46. This is the exact same as our treatment coefficient of 2.46 in Q4.4 using the linear model in 4.4. The model predicts that the treatment led to a 2.46 point increase in the follow-up attitude thermometer score.
